Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-17 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Last Revision: small changes to improve readibility

\medskip
As demonstrated by \citet{habermann_1960} \citep[see also][]{budek_2017}, mobility ratio (MR) is a key-parameter to assess fluid displacement and is defined as the ratio of mobility of the displacing (fluid $i$) to that of the displaced fluid ($j$),
\begin{equation}
 \text{MR} \; = \frac{\mathcal{K}_{ri} \mu_{j}}{\mathcal{K}_{rj} \mu_{i}}. 
\label{eq:MR}
\end{equation}
MR is a function not only of fluids' viscosity, but also of the parameterised relative permeability, $\mathcal{K}_{r\alpha}$, which is often expressed as a function of local, residual and maximum phase saturations prescribed in the pore rock matrix. In the simulations conducted in Section~\ref{section:results}, the modified \cite{Brooks_1964} model was used \citep{alpak_1999},
\begin{eqnarray}
  \mathcal{K}_{rw}\left(S_{w}\right) &=& \mathcal{K}^{\circ}_{rw}\left[\frc{S_{w}-S_{w,irr}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{w}}, \label{Eqn:CoreyBrooks1}\\
  \mathcal{K}_{rnw}\left(S_{nw}\right) &=& \mathcal{K}^{\circ}_{rnw}\left[\frc{S_{nw}-S_{nw,r}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{nw}}, \label{Eqn:CoreyBrooks2}
\end{eqnarray}
where subscripts $w$ and $nw$ stand for wetting and non-wetting phases, respectively. $\mathcal{K}^{\circ}_{rw}$ and $\mathcal{K}^{\circ}_{rnw}$ are end-point relative permeability to wetting and non-wetting phases. $S_{w,irr}$ and $S_{nw,r}$ are irreducible wetting and residual non-wetting phase saturations, respectively. Exponents $n_{w}$ and $n_{nw}$ are both set to 2. 
%The MR contains the phase saturation (see Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}) while it is clear that during fluid displacement phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, will change in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
%\noindent where subscripts $w$ and $nw$ stand for wetting and non-wetting phases, respectively. $\mathcal{K}^{\circ}_{rw}$ and $\mathcal{K}^{\circ}_{rnw}$ are end-point relative permeability to wetting and non-wetting phases, $S_{w,irr}$ and $S_{nw,r}$ are irreducible wetting and residual non-wetting phase saturations, respectively.

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqns.~\ref{Eqn:CoreyBrooks1}-\ref{Eqn:CoreyBrooks2}), however it is clear that during fluid displacement they change in time and space according to the global mass conservation equation. Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
which will be used in the parametrisation of the numerical simulations conducted in Section~\ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$.

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model's capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The discontinuous spatial permeability distribution (across the regions) leads to preferential flow through more permeable regions $\left(\text{\ie through K}_{1}\right)$. %Figure \ref{fem_cv_represent_a}(b-c)describe the saturation field at different times.

%A numerical simulation was also performed in modified conditions (see Fig.~\ref{fem_cv_represent_a}) to assess model behaviour under larger {\it VR}=10. Initial and boundary conditions remain the same, and flow behaviour qualitatively agree validated against laboratory experiments conducted by \citet{evans_1994}. 
A two-phase immiscible flow along homogeneous and parallel layers of contrasting petrophysical properties (\ie permeability) is initially simulated. During fluid displacement, crossflow between adjacent layers often occur due to viscous, capillary and/or gravitational forces that drive the flow. In this numerical simulation, crossflow is caused by viscous forces, which is commonly named viscous crossflow. Figures~\ref{fem_cv_represent_a}(b) and (c) show continuous displacement of fluid 2 due to the injection of fluid 1. They also demonstrate the preferential flow path through high-permeability regions. Such flow behaviour, represented by the crossflow through first and fourth quadrant of the domain, is in good qualitative agreement with experiments conducted by \citet{dawe_2008} (see Figs. 5 and 6 of this reference).

>>>>

\medskip
As demonstrated by \citet{habermann_1960} \citep[see also][]{budek_2017}, mobility ratio (MR) is a key-parameter to assess fluid displacement and is defined as the ratio of mobility of the displacing (fluid $i$) to that of the displaced fluid ($j$),
\begin{equation}
 \text{MR} \; = \frac{\mathcal{K}_{ri} \mu_{j}}{\mathcal{K}_{rj} \mu_{i}}. 
\label{eq:MR}
\end{equation}
MR is a function not only of fluids' viscosity, but also of the parameterised relative permeability, $\mathcal{K}_{r\alpha}$, which is often expressed as a function of local, residual and maximum phase saturations prescribed in the pore rock matrix. In the simulations conducted in Section~\ref{section:results}, the modified \cite{Brooks_1964} model was used \citep{alpak_1999},
\begin{equation}
   \begin{cases}
      \mathcal{K}_{rw}\left(S_{w}\right) = \mathcal{K}^{\circ}_{rw}\left[\frc{S_{w}-S_{w,irr}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{w}}, & \\%\label{Eqn:CoreyBrooks1}\\
                                                                                                                         & \\
      \mathcal{K}_{rnw}\left(S_{nw}\right) = \mathcal{K}^{\circ}_{rnw}\left[\frc{S_{nw}-S_{nw,r}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{nw}}, & \label{Eqn:CoreyBrooks}%\label{Eqn:CoreyBrooks2}
   \end{cases}
\end{equation}
where subscripts $w$ and $nw$ stand for wetting and non-wetting phases, respectively. $\mathcal{K}^{\circ}_{rw}$ and $\mathcal{K}^{\circ}_{rnw}$ are end-point relative permeability to wetting and non-wetting phases. $S_{w,irr}$ and $S_{nw,r}$ are irreducible wetting and residual non-wetting phase saturations, respectively. Exponents $n_{w}$ and $n_{nw}$ are both set to 2. 
%The MR contains the phase saturation (see Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}) while it is clear that during fluid displacement phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, will change in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
%\noindent where subscripts $w$ and $nw$ stand for wetting and non-wetting phases, respectively. $\mathcal{K}^{\circ}_{rw}$ and $\mathcal{K}^{\circ}_{rnw}$ are end-point relative permeability to wetting and non-wetting phases, $S_{w,irr}$ and $S_{nw,r}$ are irreducible wetting and residual non-wetting phase saturations, respectively.

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqn.~\ref{Eqn:CoreyBrooks}), however it is clear that during fluid displacement they change in time and space according to the global mass conservation equation. Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
which will be used in the parametrisation of the numerical simulations conducted in Section~\ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$.

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model's capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u_{x}=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a).

A two-phase immiscible flow along homogeneous and parallel layers of contrasting petrophysical properties (\ie permeability) is initially simulated. During fluid displacement, crossflow between adjacent layers often occur due to viscous, capillary and/or gravitational forces in discontinuous spatial permeability distribution. Such preferential flows are shown in Figs.~\ref{fem_cv_represent_a}(b-c). In this case, the crossflow dynamics through first and fourth quadrants of the domain is in good qualitative agreement with experiments conducted by \citet{dawe_2008} (see Figs. 5 and 6 of this reference).

%A numerical simulation was also performed in modified conditions (see Fig.~\ref{fem_cv_represent_a}) to assess model behaviour under larger {\it VR}=10. Initial and boundary conditions remain the same, and flow behaviour qualitatively agree validated against laboratory experiments conducted by \citet{evans_1994}. 
%XXXXXXXXXXXXXXXXXXXXX
%A two-phase immiscible flow along homogeneous and parallel layers of contrasting petrophysical properties (\ie permeability) is initially simulated. During fluid displacement, crossflow between adjacent layers often occur due to viscous, capillary and/or gravitational forces that drive the flow. In this numerical simulation, crossflow is caused by viscous forces, which is commonly named viscous crossflow. Figures~\ref{fem_cv_represent_a}(b) and (c) show continuous displacement of fluid 2 due to the injection of fluid 1. They also demonstrate the preferential flow path through high-permeability regions. Such flow behaviour, represented by the crossflow through first and fourth quadrant of the domain, is in good qualitative agreement with experiments conducted by \citet{dawe_2008} (see Figs. 5 and 6 of this reference).

### 2018-07-17 Papers/ViscousInstabilities/article_table.tex
updated
update
update the numbers - not look like a code
updates
Last Revision: small changes to improve readibility

%%%
%%%   TABLE 1
%%%
\begin{landscape}
\begin{table}
  \begin{tabular}{c | c c  c  c  c  c  c  c  c  c   c}
    \hline
      {\bf Section} & $\phi$ & VR  & $S^{0}_{w}$ & $S^{0}_{nw}$ & $\mathbf{K}_{1}$ & $\mathbf{K}_{2}$ & $\mathbf{K}_{3}$ & $\mathbf{K}_{4}$ & $S_{w,irr}$ & $S_{nw,r}$ & $u^{\circ}_{w}$ \\ 
    \hline
     \ref{section:results_validation} & 0.2 & 1.0 & 0.0 & 1.0 & 2.5 & 1.0 & N/A & N/A & 0.2 & 0.3 & 1.0 \\
     \ref{section:results_homo_hete}(homogenous)  & 0.2 & 10.0 & 0.0 & 1.0 & 1.e-10 & N/A & N/A & N/A & 0.2 & 0.3 & 0.5  \\
     \ref{section:results_homo_hete}(heterogenous) & 0.2 & 1.0 & 0.0 & 1.0 & 1.e-11 -- & 1.e-12 -- & 1.e-12 -- & 1.e-10 & 0.2  & 0.3 & 1.0 \\
      &   &  &  &  & 5.e-10 & 5.e-10 & 1.e-10  &  &  &  & \\
     \ref{section:results_hete_fix_adapt}  & 0.2 & 1.0 & 0.0 & 1.0  & 3.0 & 2.0 & 5.0 & 1.0 & 0.2 & 0.3 & 0.5  \\
     \ref{section:results_dinlet}  & 0.2 & 1.0 & 0.0 & 1.0 & 3.0 & 2.0 & 5.0 & 1.0 & 0.2  & 0.3 & 0.5  \\
     \hline
   \end{tabular}
   \caption{Sumary of model set-up used in the numerical simulations. Superscript $\circ$ denotes initial condition. $\mathbf{K}_{i}$ is in $m^{2}$ and $u^{\circ}_{w}$ is in m/sec (units). $S_{w,irr}$ and $S_{nw,r}$ are the same for all simulations. }
\label{table:setup}
\end{table}
\end{landscape}
\clearpage

%%%
%%%   TABLE 2
%%%
\begin{table}
  \begin{tabular}{c | c c c c}
    \hline\hline
                                  & {\bf K}                                  & {\it VR}  & $\lambda_{f}^{\text{(calc,max)}}$ (cm) & $\lambda_{f}^{\text{(simul,max)}}$ (cm) \\
                                  & $\left(\text{cm}^{2}\right)$              &           &  (Eqn.~\ref{eqn:guaneqn})  &                    \\
    \hline\hline
    {\bf Case 1} (homogeneous)    &  1$\times$10$^{-10}$                      &    10      &  0.13-0.75                 &   0.45-0.70       \\
    {\bf Case 2} (homogeneous)    &  1$\times$10$^{-10}$                      &    150     &  0.30-0.75                 &   0.50-0.90       \\
    {\bf Case 3} (heterogeneous)  &  1$\times$10$^{-12}$-5$\times$10$^{-10}$   &    10      &  0.25-4.50                 &   0.44-0.56       \\
    {\bf Case 4} (heterogeneous)  &  1$\times$10$^{-12}$-5$\times$10$^{-10}$   &    150     &  0.88-3.00                 &   0.24-0.40       
  \end{tabular}
   \caption{Flow simulations in Heleshaw cells (Section~\ref{section:results_homo_hete}): summary of finger width calculations using Eqn.~\ref{eqn:guaneqn} and obtained through numerical simualtions. $\lambda_{f}^{\text{calc}}$ was calculated based on fingers' tip velocities and outflow rate. Calculations were based on simulation data obtained from snapshots shown in Fig.~\ref{fig:homoheleshaw_VN10_VN150}.}\label{table:Heleshaw}
\end{table}

>>>>

%%%
%%%   TABLE 1
%%%
\begin{landscape}
\begin{table}
  \begin{tabular}{c | c c  c  c  c  c  c  c  c  c   c}
    \hline
      {\bf Section} & $\phi$ & VR  & $S^{0}_{w}$ & $S^{0}_{nw}$ & $\mathbf{K}_{1}$ & $\mathbf{K}_{2}$ & $\mathbf{K}_{3}$ & $\mathbf{K}_{4}$ & $S_{w,irr}$ & $S_{nw,r}$ & $u^{\circ}_{w}$ \\ 
    \hline
     \ref{section:results_initial_model_validation} & 0.2 & 1 & 0.0 & 1.0 & 2.5 & 1.0 & -- & -- & 0.2 & 0.3 & 1.0 \\
     \ref{section:results_homo_hete}(homogenous)  & 0.2 & 3/10/150 & 0.0 & 1.0 & 1$\times$10$^{-10}$ & -- & -- & -- & 0.2 & 0.3 & 1.0  \\
     %                                             &     & 150 &     &     &                    &     &     &     &     &     &      \\
     \ref{section:results_homo_hete}(heterogenous) & 0.2 & 10/150 & 0.0 & 1.0 & 1$\times$10$^{-11} $-- & 1$\times$10$^{-12}$ -- & 1$\times$10$^{-12}$ -- & 1$\times$10$^{-10}$ & 0.2  & 0.3 & 1.0 \\
      &   &  &  &  & 5$\times$10$^{-10}$ & 5$\times$10$^{-10}$ & 1$\times$10$^{-10}$  &  &  &  & \\
     \ref{section:results_hete_fix_adapt}  & 0.2 & 10 & 0.0 & 1.0  & 3.0 & 2.0 & 5.0 & 1.0 & 0.2 & 0.3 & 0.5  \\
     \ref{section:results_3D}  & 0.2 & 150 & 0.0 & 1.0 & 3.0 & 2.0 & 5.0 & 1.0 & 0.2  & 0.3 & 0.5  \\
     \hline
   \end{tabular}
   \caption{Sumary of model set-up used in the numerical simulations. Superscript $\circ$ denotes initial condition. $\mathbf{K}_{i}$ is in cm$^{2}$ and $u^{\circ}_{w}$ is in cm.s$^{-1}$ (except for Sections~\ref{section:results_initial_model_validation} and \ref{section:results_hete_fix_adapt} with dimensionless units). $S_{w,irr}$ and $S_{nw,r}$ are the same for all simulations (Eqn.~\ref{Eqn:CoreyBrooks}).}
\label{table:setup}
\end{table}
\end{landscape}
\clearpage

%%%
%%%   TABLE 2
%%%
\begin{landscape}
\begin{table}
  \begin{tabular}{c | c c c c}
    \hline\hline
                                  & {\bf K}                                  & {\it VR}  & $\lambda_{f}^{\text{(calc,max)}}$ (cm) & $\lambda_{f}^{\text{(simul,max)}}$ (cm) \\
                                  & $\left(\text{cm}^{2}\right)$              &           &  (Eqn.~\ref{eqn:guaneqn})  &                    \\
    \hline\hline
    {\bf Case 1} (homogeneous)    &  1$\times$10$^{-10}$                      &    10      &  0.13-0.75                 &   0.45-0.70       \\
    {\bf Case 2} (homogeneous)    &  1$\times$10$^{-10}$                      &    150     &  0.30-0.75                 &   0.50-0.90       \\
    {\bf Case 3} (heterogeneous)  &  1$\times$10$^{-12}$-5$\times$10$^{-10}$   &    10      &  0.25-4.50                 &   0.44-0.56       \\
    {\bf Case 4} (heterogeneous)  &  1$\times$10$^{-12}$-5$\times$10$^{-10}$   &    150     &  0.88-3.00                 &   0.24-0.40       
  \end{tabular}
   \caption{Flow simulations in Heleshaw cells (Section~\ref{section:results_homo_hete}): summary of finger width calculations using Eqn.~\ref{eqn:guaneqn} and obtained through numerical simualtions. $\lambda_{f}^{\text{calc}}$ was calculated based on fingers' tip velocities and outflow rate. Calculations were based on simulation data obtained from snapshots shown in Fig.~\ref{fig:homoheleshaw_VN10_VN150}.}\label{table:Heleshaw}
\end{table}
\end{landscape}
\clearpage
